\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2019-06-21 IPANEMAP-2019.tex
Updates from Overleaf

SHAPE reagents represent a popular class of small molecules that react with the 2’ hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As Ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents to discriminate nucleotides in stable interactions from others. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamics tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates N1 of adenines and N3 of cytosine if they are not involved in a hydrogen bond. Similarly, 1-cyclohexyl-3-(2-morpholinoethyl)carbodiimide metho-p-toluene sulfonate (CMCT) reacts on the Watson-Crick face of guanosine (N1) and uracyl (N3) when not engaged in a hydrogen bond~\cite{Ehresmann1987,Brunel2000}. These two reagents therefore mostly reveal Watson-Crick base pairing, but also any other type of contact involving the same edge. This diversity of probes not only increases  coverage of the different positions and structural contexts in an RNA, but it also provides qualitatively different qualitative information, as been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017}.

@@

\paragraph{Hajdin \emph{et al} dataset.} 
A dataset was gathered by Hajdin\etal\cite{Hajdin2013} to validate the predictive capacities of probing data-driven predictions. It consists of $24$ RNA sequences with known secondary structures for which a single chemical probing reactivity profile (SHAPE -- 1M7) is available.
This dataset includes sequences originating from a variety of organisms, and spans lengths ranging from $34$ nts to $530$ nts, with a focus on riboswitches and complex RNA architectures.

@@

\paragraph{Stops-inducing probing protocol (SHAPE-CE).} 
6 pmol of RNA were resuspended in 18 $\mu$l of water, denatured at 80\degree{}C for 2 minutes and cooled down at room temperature during 10 minutes in the probing buffer (40 mM HEPES at 7.5 pH, 100 mM KCl, in presence or absence of 5 mM MgCl$_\text{2}$. After a 10 minutes incubation at 37\degree{}C, RNAs were treated with 2 mM of SHAPE reagent or DMSO (negative control) and incubated for 2 (BzCN), 5 (1M7), 30 (NMIA) or 60 (NAI) minutes at 37\degree{}C. Modified or unmodified RNAs were purified by ethanol precipitation and pellets were resuspended in 10 $\mu$l of water.
Modifications were revealed by reverse transcription using 5’ fluorescently labelled primers (D2 or D4 WellRED, Sigma Aldrich 5’ CTG-TGA-ACT-AAT-GCT-GTC-CTT-TAA 3’) and M-MLV RNAse H- reverse transcriptase (Promega) as previously described~\cite{Deforges2017}, with only minor modification of the originally described SHAPE protocol~\cite{Wilkinson2006}. cDNAs were separated by capillary electrophoresis (Beckman Coulter, Ceq8000). Data were analyzed using the  \Software{QuSHAPE}~\cite{Karabiber2013} software. RNA probing was performed in triplicate with distinct RNA preparations. The corresponding data sets are named after the probe followed by "-CE" and Mg if probed in presence of MgCl$_\text{2}$.

@@

For 1M7ILU3, the denaturation step was repeated three times : After a 1 minute incubation at 95\degree{}C 10 mM of 1M7 was added and the mix was incubated 1 minute at 95\degree{}C. This step was repeated 3 times(3D). For 1M7, 1M7Mg, NMIA and NMIAMg, experiments were conducted as described above, except that the NGS library preparation was adapted to Ion Torrent sequencing. Sequences were mapped and analyzed with \Software{ShapeMapper2} for 1M7ILU and 1M7ILU3. For the other conditions they are mapped with a script based on \Software{ShapeMapper}, but adapted to Ion Torrent output files.The corresponding data sets are named after the probe followed by "-MAP", Mg if probed in presence of MgCl$_\text{2}$, and IL  or IT for Illumina or Ion Torrent sequencing, respectively.

\paragraph{DiMethylSulfate (DMS) and CMCT probing.}
DMS and CMCT probing were conducted essentially as described in~\cite{Weill2004,James2008}. Succinctly, 6 pmol of RNA were resuspended in 18 $\mu$l of water, denatured at 80\degree{}C for 2 minutes and cooled down at room temperature for 10 minutes in a probing buffer for DMS probing (40 mM HEPES pH 7.5, 100 mM KCl, 5 mM MgCl$_\text{2}$) or in CMCT (50 mM of potassium borate pH 8, 100 mM KCl and 5 mM MgCl$_\text{2}$). For DMS, RNA was then treated with 1 $\mu$l of DMS (1:12 in ethanol) or 1 $\mu$l of ethanol for 5 minutes at 37\degree{}C (mock reaction), the reaction was stopped by addition of 400 mM of Tris at 7.5 pH and immediately put on ice. For CMCT,  RNA was treated with 10 $\mu$l of CMCT (42mg/mL) or 10 $\mu$l H$_\text{2}$O for 10 minutes at 37\degree{}C (mock reaction). Modified RNA were ethanol-precipitated and resuspended in 10 $\mu$l of water. Modifications were mapped as described above for SHAPE-CE experiments.The corresponding data sets are named after the probe followed by "-CE" and Mg.

@@

Interestingly, in the absence of probing data, MFE predictions are generally dominant on this dataset, trailed by the MEA, and followed by \OurTool which, in this setting, devolves into the Ding-Lawrence algorithm~\citep{Ding2003}. However, whenever probing reactivities are available, the single centroid returned by \OurTool always achieves higher GM values (Avg: $70\%$) than both MEA (Avg: $62\%$) and MFE  (Avg: $58\%$), whose relative performances depend on the probing reagent.%la moyenne sur les 6 ARNS et sur les 3 reactifs
Interestingly, the quality of MFE and MEA predictions does not systematically benefit from additional probing data. Indeed, for half of the reagents and methods, the average GM obtained in the presence of a single reagent is lower than in the absence of probing data. Also, the impact of single probing data varies greatly across the three reagents, and the GE values of predictions respectively informed by CMCT, DMS and NMIA are ordered increasingly for all approaches, except for a minor reversal of DMS and NMIA in MEA mode.

@@

\subsection*{\OurTool outperforms state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assess the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compare its performances against \Software{RSample} software recently introduced by Spacic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform favorably against a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAstructure} suite~\citep{Reuter2010}.
We consider the dataset of hajdin~\emph{et al}, which consists of 24 RNAs of lengths ranging from 47 to 500 nucleotides, all believed to fold into a unique documented conformation.

@@

	\caption{Model of the Lariat capping ribozyme from {\itshape Didymium iridis} (A -- PDB 4P8Z:A by Meyer\etal\cite{Meyer2014}), secondary structure, as annotated by \Software{DSSR} (B -- pseudoknot ommitted) and \OurTool first-ranking prediction (C.), obtained from the integration of 8 maximally-diverging probing conditions (\NMIAMg, \OneMSevILUThreeMg, \NAIMg, \OneMSevCE, \CMCTMg,\OneMSevILUThree, and \OneMSevMgCE).\label{fig:didy}}
\end{figure*}
As benchmark RNA we used the lariat capping ribozyme from \textit{Didymium Iridis} (DiLCrz) which 3D structure has been modeled at 3,5 A from X-ray diffraction data~\cite{Meyer2014}. This RNA contains a large diversity of interactions, including helical domains, tertiary pairings, a kissing loop and a pseudoknot (Figure~\ref{fig:didy}).
It was probed as described in material and methods in 16 different conditions using different SHAPE and classical reagents, in presence/absence of Mg$^\text{2+}$ ions, using either the standard premature stop or the newly developed mutation induction technologies to map the modification sites. These experimental results were then used as input of \OurTool{}, either individually or combined,  for secondary structure prediction.

\paragraph{Comparison and clustering of conditions.}
Faced with diverse probing conditions, we first assess the compatibility of the conclusions drawn from different probing data, including the probing-free run of \OurTool as a control. We use the methodology described in the material and methods~\ref{sec:dotplots} and, for each condition, compute the base pair probability distribution (aka \emph{dot-plot}) in the  pseudo-Boltzmann ensemble, induced by the reactivity profile. We then computed the ensemble distance, the squared Euclidean distance between dot plots, for each pair of conditions. Figure~\ref{fig:PCA} summarizes the pairwise distance, projected onto a 2D surface by a principal component analysis.

>>>>

SHAPE reagents represent a popular class of small molecules that react with the 2’ hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents to discriminate nucleotides in stable interactions from others. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamics tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates N1 of adenines and N3 of cytosine if they are not involved in a hydrogen bond. Similarly, 1-cyclohexyl-3-(2-morpholinoethyl)carbodiimide metho-p-toluene sulfonate (CMCT) reacts on the Watson-Crick face of guanosine (N1) and uracyl (N3) when not engaged in a hydrogen bond~\cite{Ehresmann1987,Brunel2000}. These two reagents therefore mostly reveal Watson-Crick base pairing, but also any other type of contact involving the same edge. This diversity of probes not only increases  coverage of the different positions and structural contexts in an RNA, but it also provides qualitatively different qualitative information, as been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017}.

@@

\paragraph{Hajdin \emph{et al} dataset.} 
A dataset was gathered by Hajdin\etal\cite{Hajdin2013} to validate the predictive capacities of probing data-driven predictions. It consists of $24$ RNA sequences with known secondary structures for which a single chemical probing reactivity profile (1M7-\SH) is available.
This dataset includes sequences originating from a variety of organisms, and spans lengths ranging from $34$ nts to $530$ nts, with a focus on riboswitches and complex RNA architectures.

@@

\paragraph{Stops-inducing probing protocol (SHAPE-CE).} 
6 pmol of RNA were resuspended in 18 $\mu$l of water, denatured at 80\degree{}C for 2 minutes and cooled down at room temperature during 10 minutes in the probing buffer (40 mM HEPES at 7.5 pH, 100 mM KCl, in presence or absence of 5 mM MgCl$_\text{2}$. After a 10 minutes incubation at 37\degree{}C, RNAs were treated with 2 mM of SHAPE reagent or DMSO (negative control) and incubated for 2 (BzCN), 5 (1M7), 30 (NMIA) or 60 (NAI) minutes at 37\degree{}C. Modified or unmodified RNAs were purified by ethanol precipitation and pellets were resuspended in 10 $\mu$l of water.
Modifications were revealed by reverse transcription using 5' fluorescently labelled primers (D2 or D4 WellRED, Sigma Aldrich 5' CTG-TGA-ACT-AAT-GCT-GTC-CTT-TAA 3') and M-MLV RNAse (H-) reverse transcriptase (Promega) as previously described~\cite{Deforges2017}, with only minor modification of the originally described SHAPE protocol~\cite{Wilkinson2006}. cDNAs were separated by capillary electrophoresis (Beckman Coulter, Ceq8000). Data were analyzed using the  \Software{QuSHAPE}~\cite{Karabiber2013} software. RNA probing was performed in triplicate with distinct RNA preparations. The corresponding data sets are named after the probe followed by "-CE" and Mg if probed in presence of MgCl$_\text{2}$.

@@

For \OneMSevILUThreeMg, the denaturation step was repeated three times : After a 1 minute incubation at 95\degree{}C, 10 mM of 1M7 was added and the mix was incubated 1 minute at 95\degree{}C. This step was repeated 3 times (3D). For 1M7, 1M7Mg, NMIA and NMIAMg, experiments were conducted as described above, except that the NGS library preparation was adapted to Ion Torrent sequencing. Sequences were mapped and analyzed with \Software{ShapeMapper2} for \OneMSevILUThreeMg and \OneMSevILUThreeMg. For the other conditions they are mapped with a script based on \Software{ShapeMapper}, but adapted to Ion Torrent output files. The corresponding data sets are named after the probe followed by "-MAP", Mg if probed in presence of MgCl$_\text{2}$, and IL  or IT for Illumina or Ion Torrent sequencing, respectively.

\paragraph{DiMethylSulfate (DMS) and CMCT probing.}
DMS and CMCT probing were conducted essentially as described in~\cite{Weill2004,James2008}. Succinctly, 6 pmol of RNA were resuspended in 18 $\mu$l of water, denatured at 80\degree{}C for 2 minutes and cooled down at room temperature for 10 minutes in a probing buffer for DMS probing (40 mM HEPES pH 7.5, 100 mM KCl, 5 mM MgCl$_\text{2}$) or in CMCT (50 mM of potassium borate pH 8, 100 mM KCl and 5 mM MgCl$_\text{2}$). For DMS, RNA was then treated with 1 $\mu$l of DMS (1:12 in ethanol) or 1 $\mu$l of ethanol for 5 minutes at 37\degree{}C (mock reaction), the reaction was stopped by addition of 400 mM of Tris at 7.5 pH and immediately put on ice. For CMCT,  RNA was treated with 10 $\mu$l of CMCT (42 mg/mL) or 10 $\mu$l H$_\text{2}$O for 10 minutes at 37\degree{}C (mock reaction). Modified RNA were ethanol-precipitated and resuspended in 10 $\mu$l of water. Modifications were mapped as described above for SHAPE-CE experiments.The corresponding data sets are named after the probe followed by "-CE" and Mg.

@@

Interestingly, in the absence of probing data, MFE predictions are generally dominant on this dataset, trailed by the MEA, and followed by \OurTool which, in this setting, devolves into the Ding-Lawrence algorithm~\citep{Ding2003}. However, whenever probing reactivities are available, the single centroid returned by \OurTool always achieves higher GM values (Avg: $70\%$) than both MEA (Avg: $62\%$) and MFE  (Avg: $58\%$), whose relative performances depend on the probing reagent. %la moyenne sur les 6 ARNS et sur les 3 reactifs
Interestingly, the quality of MFE and MEA predictions does not systematically benefit from additional probing data. Indeed, for half of the reagents and methods, the average GM obtained in the presence of a single reagent is lower than in the absence of probing data. Also, the impact of single probing data varies greatly across the three reagents, and the GE values of predictions respectively informed by CMCT, DMS and NMIA are ordered increasingly for all approaches, except for a minor reversal of DMS and NMIA in MEA mode.

@@

\subsection*{\OurTool outperforms state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assess the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compare its performances against \Software{RSample} software recently introduced by Spacic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform favorably against a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAStructure} suite~\citep{Reuter2010}.
We consider the dataset of hajdin~\emph{et al}, which consists of 24 RNAs of lengths ranging from 47 to 500 nucleotides, all believed to fold into a unique documented conformation.

@@

	\caption{Model of the Lariat capping ribozyme from {\itshape Didymium iridis} (A -- PDB 4P8Z:A by Meyer\etal\cite{Meyer2014}), secondary structure, as annotated by \Software{DSSR} (B -- pseudoknot ommitted) and \OurTool first-ranking prediction (C.), obtained from the integration of 8 maximally-diverging probing conditions (\NMIAMg, \OneMSevILUThreeMg, \NAIMg, \OneMSevCE, \CMCTMg,\OneMSevILUThree, and \OneMSevMgCE).\label{fig:didy}}
\end{figure*}
As benchmark RNA we used the Lariat capping ribozyme from \textit{Didymium Iridis} (DiLCrz) which 3D structure has been modeled at 3,85 A from X-ray diffraction data~\cite{Meyer2014}. This RNA contains a large diversity of interactions, including helical domains, tertiary pairings, a kissing loop and a pseudoknot (Figure~\ref{fig:didy}).
It was probed as described in Material and Methods in 16 different conditions using different SHAPE and classical reagents, in presence/absence of Mg$^\text{2+}$ ions, using either the standard premature stop or the newly developed mutation induction technologies to map the modification sites. These experimental results were then used as input of \OurTool{}, either individually or combined,  for secondary structure prediction.

\paragraph{Comparison and clustering of conditions.}
Faced with diverse probing conditions, we first assess the compatibility of the conclusions drawn from different probing data, including the probing-free run of \OurTool as a control. We use the methodology described in the Material and Methods~\ref{sec:dotplots} and, for each condition, compute the base pair probability distribution (aka \emph{dot-plot}) in the  pseudo-Boltzmann ensemble, induced by the reactivity profile. We then computed the ensemble distance, the squared Euclidean distance between dot plots, for each pair of conditions. Figure~\ref{fig:PCA} summarizes the pairwise distance, projected onto a 2D surface by a principal component analysis.
